---
title: "Title Express Data Extract & PRep"
author: "Colin Lacy - syntaxcritic"
output:
  html_document: 
    code_folding: hide
    df_print: paged
---


# 1 - Load Packages

Load packages for data extraction and prep.

```{r Load Packages}
library(tidyverse)
library(readxl)
library(lubridate)
library(knitr)

```



# 2 - Choose Files for Analysis

Define locations of data tables exported from SMS Title Express.

```{r Set filepaths for TSS data}
# Locate "Dates" data
file.choose()
filepath_dates <- .Last.value

# Locate "Address" data
file.choose()
filepath_address <- .Last.value

# Locate "Loan and Lender" data
file.choose()
filepath_loan_lender <- .Last.value

# Locate "Sale and Loan Amounts" data
file.choose()
filepath_sale_loan_amts <- .Last.value

# Locate "Escrow Account" data
file.choose()
filepath_escrow <- .Last.value

# Locate "Policy Log" data
file.choose()
filepath_policy_log <- .Last.value


```



# 3 - Extract "Dates" data

**Key Inputs:** 
* filepath_dates

**Key Outputs:** 
* dates_data

**Steps:**
* read .csv data from "Dates" file 
* define variables
* convert dates from M/D/YYYY to YYYY-MM-DD format
* clean "Order Number" variable (upper-case, no spaces)

```{r Extract Dates data}
dates_data <- read_csv(filepath_dates, 
                       col_types = cols(
                         `Order Number` = col_character(), 
                         `Settlement Date` = col_date(format = "%m/%d/%Y"), 
                         `Disbursement Date` = col_date(format = "%m/%d/%Y"), 
                         `Closed Date` = col_date(format = "%m/%d/%Y"), 
                         `Open Date` = col_date(format = "%m/%d/%Y"), 
                         `Lender Policy Date` = col_skip(), 
                         `Owners Policy Date` = col_skip(), 
                         `Lender's Distribution Date` = col_skip(),
                         `Owner's Distribution Date` = col_skip(), 
                         `Issued Date` = col_skip(), 
                         `Confirmed Date` = col_skip(),
                         `Buyer 1` = col_character(), 
                         `Seller 1` = col_character(), 
                         `Client Name` = col_character(), 
                         `Lender Name` = col_character(), 
                         `Officer` = col_skip(), 
                         `Closer` = col_skip(), 
                         `Status` = col_character(), 
                         `Address 1` = col_character(), 
                         `City` = col_character(), 
                         `County` = col_character(), 
                         `State` = col_character(), 
                         `Branch` = col_skip() )) %>% 
  mutate(`Order Number` = str_to_upper(`Order Number`)) %>% 
  mutate(`Order Number` = str_remove_all(`Order Number`, "\\s"))

```



# 4 - Extract Addresses

**Key Inputs:**
* filepath_address

**Key Outputs:**
* address_data

**Steps:**
* read .csv data from "Address" file 
* define variables
* clean "Order Number" variable (upper-case, no spaces)

```{r Extract Address data}
address_data <- read_csv(filepath_address, 
                         col_types = cols(
                           `Order Number` = col_character(),
                           `Address 1` = col_character(), 
                           `Address 2` = col_character(), 
                           `City` = col_character(), 
                           `County` = col_character(), 
                           `State` = col_character(), 
                           `ZIP` = col_integer(), 
                           `Buyer 1` = col_character(), 
                           `Seller 1` = col_character(), 
                           `Officer` = col_skip(), 
                           `Closer` = col_skip(), 
                           `Status` = col_character(), 
                           `Branch` = col_skip() )) %>% 
  mutate(`Order Number` = str_to_upper(`Order Number`)) %>% 
  mutate(`Order Number` = str_remove_all(`Order Number`, "\\s"))

```



# 5 - Extract Lenders

**Key Inputs:**
* filepath_loan

**Key Outputs:**
loan_lender_data

**Steps:**
* read .csv data from "Loan & Lender" file 
* define variables
* clean "Order Number" variable (upper-case, no spaces)

```{r Extract "Find, Loan" for DC/MD/VA}
loan_DC <- read_csv(filepath_loan_DC, 
                    col_types = cols(
                      `Order Number` = col_character(), 
                      `Loan Number` = col_number(), 
                      `Lender Name` = col_character(), 
                      `Lender Agent` = col_skip(), 
                      `Buyer 1` = col_character(), 
                      `Seller 1` = col_character(), 
                      `AU Case File ID` = col_skip(), 
                      `UCD Case File ID` = col_skip(), 
                      `Officer` = col_skip(), 
                      `Closer` = col_skip(), 
                      `Status` = col_character(), 
                      `Address 1` = col_character(), 
                      `City` = col_character(), 
                      `County` = col_character(), 
                      `State` = col_character(), 
                      `Branch` = col_skip() )) %>% 
  mutate(`Order Number` = str_to_upper(`Order Number`)) %>% 
  mutate(`Order Number` = str_remove_all(`Order Number`, "\\s"))

```



# 6 - Extract Sale and Loan Amounts

**Key Inputs:**
* filepath_sale_loan_amts

**Key Outputs:**
* sale_loan_amts

**Steps:**
* read .xlsx data from "Sale & Loan Amounts" file
* trim header/footer rows 
* define variables 
* convert date from M/D/YYYY to YYYY-MM-DD format
* clean "Order Number" variable (upper-case, no spaces)

```{r Extract "Management, History, Open Date"}
sale_loan_amts <- read_excel(filepath_sale_loan_amts, 
                                  skip = 5, 
                                  n_max = (
                                    nrow(read_excel(filepath_sale_loan_amts)
                                    )-8), 
                                  col_names = c(
                                    "Order Number", 
                                    "Type", 
                                    "Status", 
                                    "Buyer Name", 
                                    "Open Date", 
                                    "Sale Amount", 
                                    "Loan Amount", 
                                    "Event"), 
                                  col_types = c("text", 
                                                "text", 
                                                "text", 
                                                "text", 
                                                "date", 
                                                "numeric", 
                                                "numeric", 
                                                "text")) %>% 
  mutate(`Open Date` = date(`Open Date`)) %>% 
  mutate(`Order Number` = str_to_upper(`Order Number`)) %>% 
  mutate(`Order Number` = str_remove_all(`Order Number`, "\\s"))

```



# 7 - Combine Basic Case Data (Dates, Addresses, Loans, Amounts)

**Key Inputs:**
* dates_data
* address_data
* loan_lender_data
* sale_loan_amts

**Key Outputs:**
* case_info

**Steps:**
* join dates_data and address_data, then loan_lender_data
* select relevant variables
* join sale_loan_amts, by "Order Number" variable
* rename "Status" and "Disbursement Date" variables
* create "Case Type" variable
* create "Property Type" variable
* clean "Lender Name" observations (title caps, remove extra spaces)

```{r Combine dates/address/loan_lender/sale_loan_amts data}
case_info <- left_join(dates_data, address_data) %>%
  left_join(loan_lender_data) %>%
  select(
    `Order Number`:`Lender Name`, 
    `Loan Number`, 
    `Status`, 
    `Address 1`, 
    `Address 2`, 
    `City`:`ZIP`) %>% 
  left_join(sale_loan_amts, by = "Order Number") %>%
  select(
    `Order Number`:`ZIP`, 
    `Sale Amount`, 
    `Loan Amount`) %>%
  rename(`Status` = `Status.x`, 
         `Open Date` = `Open Date.x`) %>% 
  mutate(`Case Type` = case_when(
    `Sale Amount` > 0 ~ "Sale", 
    `Sale Amount` == 0 & `Loan Amount` > 0 ~ "Refi")) %>%
  mutate(`Property Type` = case_when(
    str_detect(`Address 2`, regex("apt|apartment|unit|#", ignore_case = TRUE)) ~ "Condo", 
    TRUE ~ "House")) %>%
  select(`Order Number`, 
         `Case Type`, 
         `Property Type`, 
         `Settlement Date`:`Loan Amount`) %>% 
  mutate(`Lender Name` = str_to_title(`Lender Name`)) %>%
  mutate(`Lender Name` = str_trim(`Lender Name`)) %>% 
  mutate(`Lender Name` = str_squish(`Lender Name`))

```



# 8 - Extract and Tidy FATICO Policy Logs


**Key Outputs:**
* FATICO_policies_types
* case_FATICO_policies

## 8-A
Extract the Policy Logs.

**Key Inputs:**
* filepath_policy_log

**Key Outputs:**
* FATICO_policies_all

**Steps:**
* read .XLS data from "FATICO, Policy Log, Policy Date, 2015-19, Complete" file
* trim header/footer rows
* define variables
* convert dates from M/D/YYYY to YYYY-MM-DD format
* clean "Order Number" variable (upper-case, no spaces)

```{r Extract FATICO Policy Logs}
FATICO_policies_all <- read_excel(filepath_policy_log, 
                                        skip = 10, 
                                        n_max = (nrow(read_excel(filepath_policy_log))-11), 
                                        col_names = c(
                                          "Account", 
                                          "Order Number",
                                          "Created By", 
                                          "Street", 
                                          "City", 
                                          "County", 
                                          "State", 
                                          "ZIP", 
                                          "Policy Number", 
                                          "Policy Name", 
                                          "Policy Amount", 
                                          "Status", 
                                          "Created Date", 
                                          "Policy Date", 
                                          "Rate Type", 
                                          "Extended", 
                                          "Policy Premium", 
                                          "Policy Commission", 
                                          "FATICO Policy Retention"), 
                                        col_types = c(
                                          "text", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "numeric", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "text", 
                                          "numeric", 
                                          "numeric", 
                                          "numeric")) %>% 
  mutate(`Created Date` = parse_date(`Created Date`, "%m/%d/%Y"), 
         `Policy Date` = parse_date(`Policy Date`, "%m/%d/%Y")) %>% 
  mutate(`Order Number` = str_to_upper(`Order Number`)) %>% 
  mutate(`Order Number` = str_remove_all(`Order Number`, "\\s"))

```

## 8-B
Create "Policy Type" variable ("owner" or "lender") based on "Policy Name" strings. 

**Key Inputs:**
* FATICO_policies_all

**Key Outputs:**
* FATICO_policies_types

**Steps:**
* create and store vector strings of all keywords for "Owner" and "Lender" policy types
* create "Policy Type" variable using case_when() and str_detect()
* clean "Order Number" variable (upper-case, no spaces)

```{r Create FATICO "Policy Type" variable}
owner_policy_names <- c("Eagle", "Owner")

lender_policy_names <- c("Loan")

FATICO_policies_types <- FATICO_policies_all %>% filter(`Status` == "Remitted") %>% 
  mutate(`Policy Type` = case_when(
    str_detect(`Policy Name`, regex(str_flatten(owner_policy_names, collapse = "|"), ignore_case = TRUE)) ~ "Owner", 
    str_detect(`Policy Name`, regex(str_flatten(lender_policy_names, collapse = "|"), ignore_case = TRUE)) ~ "Lender")) %>% 
  mutate(`Order Number` = str_to_upper(`Order Number`))

```

## 8-C
Create table of tidy Title Policy data (one line/observation per "Order Number").

**Key Inputs:**
* FATICO_policies_types

**Key Outputs:**
* case_FATICO_policies

**Steps:**
* select relevant variables
* group_by() key variables
* summarize(), keeping NAs
* pivot_wider() with new names from "Policy Type" variable, and rename variables
* create new variables (total Policy Premiums/Commissions/Retentions) and select key variables

```{r Tidy FATICO Policy data: one line per "Order Number"}
case_FATICO_policies <- FATICO_policies_types %>% 
  select(
    `Order Number`, 
    `State`, 
    `Policy Type`, 
    `Created Date`, 
    `Policy Date`, 
    `Policy Amount`, 
    `Policy Premium`, 
    `Policy Commission`, 
    `FATICO Policy Retention`) %>% 
  group_by(`Order Number`, 
           `Policy Type`, 
           `State`, 
           `Created Date`, 
           `Policy Date`) %>% 
  summarize(`Policy Amount` = sum(`Policy Amount`, na.rm = TRUE), 
            `Policy Premium` = sum(`Policy Premium`, na.rm = TRUE), 
            `Policy Commission` = sum(`Policy Commission`, na.rm = TRUE),
            `FATICO Policy Retention` = sum(`FATICO Policy Retention`, na.rm = TRUE)) %>% 
  pivot_wider(names_from = `Policy Type`, 
              values_from = c(`Created Date`, 
                              `Policy Date`, 
                              `Policy Amount`, 
                              `Policy Premium`, 
                              `Policy Commission`, 
                              `FATICO Policy Retention`)) %>% 
  rename(
    `Lender Policy Date` = `Policy Date_Lender`, 
    `Lender Policy Created Date` =  `Created Date_Lender`, 
    `Lender Policy Amount` = `Policy Amount_Lender`, 
    `Lender Policy Premium` = `Policy Premium_Lender`, 
    `Lender Policy Commission` = `Policy Commission_Lender`, 
    `Lender Policy FATICO Retention` = `FATICO Policy Retention_Lender`, 
    `Owner Policy Date` = `Created Date_Owner`, 
    `Owner Policy Created Date` = `Policy Date_Owner`, 
    `Owner Policy Amount` = `Policy Amount_Owner`, 
    `Owner Policy Premium` = `Policy Premium_Owner`, 
    `Owner Policy Commission` = `Policy Commission_Owner`, 
    `Owner Policy FATICO Retention` = `FATICO Policy Retention_Owner`) %>% 
  mutate(`Total Policy Premiums` = sum(`Lender Policy Premium`, 
                                      `Owner Policy Premium`, 
                                      na.rm = TRUE), 
         `Total Commissions` = sum(`Lender Policy Commission`, 
                                       `Owner Policy Commission`, 
                                       na.rm = TRUE), 
         `Total FATICO Retentions` = sum(`Lender Policy FATICO Retention`, 
                                         `Owner Policy FATICO Retention`,
                                         na.rm = TRUE)
  ) %>% 
  select(`Order Number`, 
         `State`, 
         starts_with("Total"),
         starts_with("Lender"), 
         starts_with("Owner")
  )

```



# 9 - Extract Escrow Account Transactions

**Key Inputs:**
* filepath_escrow

**Key Outputs:**
* escrow_data

**Steps:**
* read .xlsx data from "Escrow" file  
* trim header rows
* convert dates from MM/DD/YYYY to YYYY-MM-DD format
* rename variables
* clean "Order Number" variable (upper-case, no spaces)
* clean "Payee" variable (remove extra whitespace)

```{r Extract Escrow data}
escrow_data <- read_excel(filepath_escrow, skip = 5) %>% 
  mutate(Date = ymd(Date)) %>% 
  rename(`Order Number` = `File Number`) %>% 
  mutate(`Order Number` = str_to_upper(`Order Number`)) %>% 
  mutate(`Order Number` = str_remove_all(`Order Number`, "\\s")) %>% 
  mutate(`Payee` = str_trim(`Payee`)) %>% 
  mutate(`Payee` = str_squish(`Payee`))

```



# 10 - Calculate Revenue per Case
Find escrow account disbursements with your company as payee to determine total revenue per case. 

**Key Inputs:**
* Your company's name, exactly as it appears in your escrow account data when your company sends or receives funds from the escrow account
* escrow_data

**Key Outputs:**
* case_disbursement

**Steps:**
* Enter your company's name exactly as it appears in your escrow data)
* group remaining transactions by "Order Number" variable, and take sum of payments to create "Disbursement" variable
  + multiply by -1 for positive revenue

```{r Filter "escrow_data" for total funds disbursed to your company per case}

co_name <- readline(prompt = "Enter your company's name, capitalized and spelled exactly as it appears on any transactions in your escrow data: ")

case_disbursement <- escrow_data %>% 
  filter(Category == "Disbursement", 
         str_detect(Payee, 
                    co_name) == FALSE) %>% 
  group_by(`Order Number`) %>% 
  summarize(`Disbursement` = sum(Total)) %>% 
  mutate(`Disbursement` = `Disbursement` * -1) %>% 
  ungroup()

```



# 11 - Create Table of Each Order Number's County
Assists when filling in missing observations for "County" variable in final table.

**Key Inputs:**
* FATICO_policies_all

**Key Outputs:**
* county_FATICO

**Steps:**
* extract "Order Number" and "County Variables"
* change variations of county names to common spellings/names

```{r Extract Order Numbers and Counties}
county_FATICO <- FATICO_policies_all %>% select(`Order Number`, County) %>% 
  rename(CountyALT = County) %>% 
  mutate(CountyALT = str_to_title(CountyALT)) %>% 
  mutate(CountyALT = str_trim(CountyALT, side = "both")) %>% 
  mutate(CountyALT = str_squish(CountyALT)) %>% 
  mutate(CountyALT = case_when(
    str_detect(CountyALT, "District Of Columbia") ~ "DC", 
    TRUE ~ CountyALT)) %>% 
  distinct()

```


# 12 - Combine Data Into Single Table, With Detail of Revenue

**Key Inputs:**
* case_info
* case_FATICO_policies
* case_disbursement
* county_FATICO

**Key Outputs:**
* case_data

**Steps:**
* join table: case_info, case_FATICO_policies, case_disbursement
* create variables for "Service/Non-Service Fees" and "Case Revenue"
  + NOTE: subtracting with a row/observation only works properly after grouping by "Order Number". (why??)
* clean out cases with bad or suspicious data:
  + "Status" variable: "Cancelled", "Deleted", "History"
  + "Case Type" variable blank
* create "Case Transaction Amount" variable
  + equals "Sale Amount" if "Case Type" variable is a Sale, "Loan Amount" if a Refi
* create "Cash Down" and "% Cash Down" variables
* select relevant variables
* create individual month/day/year/quarter variables for each "Date" variable
* join table: county_FATICO
  + then, use its "CountyALT" to fill missing instances of "County"
  + select() out "CountyALT" variable


```{r Final Table for Your Company's Analysis & Graphs}
case_data <- left_join(case_info, case_FATICO_policies) %>% 
  left_join(case_disbursement) %>% 
  mutate(`Non-Service Fees` = sum(`Total Commissions`, na.rm = TRUE)) %>%
  mutate(`Service Fees` = `Disbursement` - `Non-Service Fees`) %>% 
  mutate(`Case Revenue` = `Disbursement`) %>% 
  ungroup() %>% 
  filter(str_detect(Status, "Cancelled|Deleted|History") == FALSE) %>% 
  filter(!is.na(`Case Type`)) %>% 
  mutate(`Case Transaction Amount` = case_when(
    `Case Type` == "Sale" ~ `Sale Amount`, 
    `Case Type` == "Refi" ~ `Loan Amount`
  )) %>% 
  mutate(`Cash Down` = case_when(
    `Case Type` == "Sale" ~ `Sale Amount` - `Loan Amount`)) %>% 
  mutate(`% Cash Down` = case_when(
    `Case Type` == "Sale" ~ round((((`Sale Amount` - `Loan Amount`)/`Sale Amount`)*100), digits = 2))) %>% 
  select(`Order Number`:`Property Type`, 
         `Open Date`, 
         `Settlement Date`:`Closed Date`, 
         `Lender Name`:`ZIP`, 
         `Case Transaction Amount`, 
         `Sale Amount`,
         `Loan Amount`, 
         `Cash Down`, 
         `% Cash Down`, 
         `Case Revenue`, 
         `Service Fees`, 
         `Total Commissions`,
         `Disbursement`, 
         `Total Policy Premiums`, 
         `Total FATICO Retentions`,
         `Buyer 1`:`Client Name`, 
         `Lender Policy Created Date`:`Owner Policy FATICO Retention`) %>% 
  rename(Lender = 'Lender Name') %>% 
  mutate(`Open Year` = year(`Open Date`), 
         `Open Month` = month(`Open Date`), 
         `Open Day` = day(`Open Date`), 
         `Open Quarter` = quarter(`Open Date`), 
         `Settlement Year` = year(`Settlement Date`), 
         `Settlement Month` = month(`Settlement Date`), 
         `Settlement Day` = day(`Settlement Date`), 
         `Settlement Quarter` = quarter(`Settlement Date`), 
         `Disbursement Year` = year(`Disbursement Date`), 
         `Disbursement Month` = month(`Disbursement Date`), 
         `Disbursement Day` = day(`Disbursement Date`), 
         `Disbursement Quarter` = quarter(`Disbursement Date`), 
         `Quarterly Open Date` = year(`Open Date`) + 0.10*(quarter(`Open Date`)), 
         `Quarterly Disbursement Date` = year(`Disbursement Date`) + 0.10*(quarter(`Disbursement Date`)), 
         `Quarterly Settlement Date` = year(`Settlement Date`) + 0.10*(quarter(`Settlement Date`))) %>% 
  left_join(county_FATICO) %>% 
  mutate(County = case_when(
    is.na(County) ~ CountyALT, 
    TRUE ~ County)) %>% 
  mutate(County = case_when(
    is.na(County) & State == "DC" ~ "DC", 
    TRUE ~ County)) %>% 
  mutate(County = str_remove(County, "City")) %>% 
  mutate(County = str_trim(County)) %>% 
  mutate(County = str_squish(County)) %>% 
  select(`Order Number`:`Quarterly Settlement Date`)

```

# 13 - Save your case data to a file

Name and save your completed file of data from Title Express and First American's policy log:

```{r}
file_name <- readline(prompt = "Enter a name for your file (no special characters): ")

write_csv(case_data, str_c(file_name, ".csv"))

```

Your new .csv file is now stored in: `r getwd()` 
